Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-04-15 Paper_Draft2/Draft2.Rmd

---
title: Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm':' A Case Study

author:
  - name: Amrita Acharya
    affil: 1, 2
  - name: Dianne Caravela
    affil: 1, 3
  - name: Eunice Kim
    affil: 1, 4
  - name: Emma Kornberg
    affil: 1, 5
  - name: Elisabeth Nesmith
    affil: 1, 6
affiliation:
  - num: 1
    address: |
      Statistical and Data Sciences
      Smith College
      Northampton, MA 01063
  - num: 2
    email: aacharya@smith.edu
  - num: 3
    email: dcaravela@smith.edu
  - num: 4
    email: ekim89@smith.edu
  - num: 5
    email: ekornberg@smith.edu
  - num: 6
    email: enesmith@smith.edu
journal: water
type: article
status: submit
bibliography: mybibfile.bib
csl: ../resources/chicago-author-date.csl
appendix: appendix.tex

@@

The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private, for-profit company Northpointe (now known by its parent company [Equivant](https://www.equivant.com/faq/)), to predict defendants’ risk of recidivism. It generates a decile score that classifies defendants’ risk of recidivism as either low, medium, or high [@angwin2016machine]. Jurisdictions across the United States use the COMPAS risk assessment instrument, including but not limited to the New York [@NYDOCCS2019], Massachusetts, Michigan, California [@Jackson2020Setting], and Wisconsin [@WisconsinDOC] Departments of Corrections. 
Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a sample COMPAS Risk Assessment Survey has been made publicly available, revealing the algorithm’s input information. This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias. For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang. It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample]. Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race. 
Angwin et al. published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine]. Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives. Angwin et al. specifically investigate the distribution of COMPAS scores by decile among Black and white defendants. They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we]. After examining the fairness metric statistical parity difference, Angwin et al. conclude that the algorithm is racially biased [@larson2016we].

@@

```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race. When we divide the data into Black and white defendants, we can see that BLack defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. We can also see that there are more Black defendants in the dataset overall.', out.width= "100%"}
knitr::include_graphics("../images/race_recid_bar_plot_new.png")
```

@@

The AIF360 toolkit contains various group and individual fairness metrics and pre-processing, in-processing, and post-processing algorithms that we used for our experiments [@aif360-oct-2018]. We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project. We chose to look at four group fairness metrics instead of individual fairness metrics. Group fairness takes into account the attributes of a whole group as opposed to just one individual in the group, allowing us to represent more of the systemic issues happening. In general, group fairness metrics require that the unprivileged group is treated similarly to the privileged group, whereas individual fairness metrics require individuals to be treated consistently [@kypraiou_what_2021]. Group and individual metrics work in opposition of one another, meaning that when group fairness improves, individual fairness gets worse.

@@

$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$

**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other. In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate. A value of 1 means that the ratio is exactly 1:1. Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits. According to AIF360, a ratio between .8 to 1.25 is considered fair:

$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$

@@

$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$

@@

$TPR_{D = Unprivileged} - TPR_{D = Privileged}$

>>>>

---
title: "Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study"

author:
  - name: Amrita Acharya
    affil: 1, 2
  - name: Dianne Caravela
    affil: 1, 3
  - name: Eunice Kim
    affil: 1, 4
  - name: Emma Kornberg
    affil: 1, 5
  - name: Elisabeth Nesmith
    affil: 1, 6
affiliation:
  - num: 1
    address: |
      Statistical and Data Sciences
      Smith College
      Northampton, MA 01063
  - num: 2
    email: aacharya@smith.edu
  - num: 3
    email: dcaravela@smith.edu
  - num: 4
    email: ekim89@smith.edu
  - num: 5
    email: ekornberg@smith.edu
  - num: 6
    email: enesmith@smith.edu
journal: water
type: article
status: submit
bibliography: mybibfile.bib
csl: ../resources/chicago-author-date.csl
#appendix: appendix.tex
classoption: fleqn

@@

The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private, for-profit company Northpointe (now known by its parent company [Equivant](https://www.equivant.com/faq/)), to predict defendants’ risk of recidivism. It generates a decile score that classifies defendants’ risk of recidivism as either low, medium, or high [@angwin2016machine]. Jurisdictions across the United States use the COMPAS risk assessment instrument, including but not limited to the New York [@NYDOCCS2019], Massachusetts, Michigan, California [@Jackson2020Setting], and Wisconsin [@WisconsinDOC] Departments of Corrections. 
Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a sample COMPAS Risk Assessment Survey has been made publicly available, revealing the algorithm’s input information. This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias. For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang. It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample]. Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race. 
Angwin et al. published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine]. Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives. Angwin et al. specifically investigate the distribution of COMPAS scores by decile among Black and white defendants. They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, Black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we]. After examining the fairness metric statistical parity difference, Angwin et al. conclude that the algorithm is racially biased [@larson2016we].

@@

```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race. When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. We can also see that there are more Black defendants in the dataset overall.', out.width= "100%"}
knitr::include_graphics("../images/race_recid_bar_plot_new.png")
```

@@

The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we used for our experiments [@aif360-oct-2018]. We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project. We chose to look at four group fairness metrics instead of individual fairness metrics. Group fairness takes into account the attributes of a whole group as opposed to just one individual in the group, allowing us to represent more of the systemic issues happening. In general, group fairness metrics require that the unprivileged group is treated similarly to the privileged group, whereas individual fairness metrics require individuals to be treated consistently [@kypraiou_what_2021]. Group and individual metrics work in opposition of one another, meaning that when group fairness improves, individual fairness gets worse.

@@

$$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$$

**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other. In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate. A value of 1 means that the ratio is exactly 1:1. Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits. According to AIF360, a ratio between .8 to 1.25 is considered fair:

$$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$$

@@

$$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$$

@@

$$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$

### 2022-04-15 Paper_Draft2/Draft2.tex

%=================================================================
% Full title of the paper (Capitalized)
\Title{Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in
the COMPAS Risk Assessment Algorithm':' A Case Study}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private, for-profit company
Northpointe (now known by its parent company
\href{https://www.equivant.com/faq/}{Equivant}), to predict defendants'
risk of recidivism. It generates a decile score that classifies
defendants' risk of recidivism as either low, medium, or high
\citep{angwin2016machine}. Jurisdictions across the United States use
the COMPAS risk assessment instrument, including but not limited to the
New York \citep{NYDOCCS2019}, Massachusetts, Michigan, California
\citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
Departments of Corrections. Due to the proprietary nature of the COMPAS
algorithm, it is unknown how exactly these recidivism risk scores are
calculated; however, a sample COMPAS Risk Assessment Survey has been
made publicly available, revealing the algorithm's input information.
This survey has been widely critiqued for using proxy variables for race
that do not explicitly factor in a defendant's race but heavily imply
it, allowing Northpointe to claim that their algorithm is free of racial
bias. For example, the COMPAS risk assessment survey asks screeners to
speculate if a defendant might be affiliated with a gang. It also asks
if a defendant has any friends or family members who have been crime
victims \citep{Angwin2016Sample}. Although these questions do not
directly ask about race, they do not take into account the pervasive
nature of systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race. Angwin et
al.~published a 2016 piece in the news publication ProPublica that
analyzes the methods and algorithms used by Northpointe in their COMPAS
risk score assessment algorithm and uncovers the racial biases in
defendants' scores \citep{angwin2016machine}. Their research finds that
``the algorithm {[}is{]} somewhat more accurate than a coin flip,'' a
worrisome level of accuracy given the potential impact its
determinations may have on real people's lives. Angwin et
al.~specifically investigate the distribution of COMPAS scores by decile
among Black and white defendants. They write: ``The analysis also
{[}shows{]} that even when controlling for prior crimes, future
recidivism, age, and gender, black defendants {[}are{]} 45 percent more
likely to be assigned higher risk scores than white defendants''
\citep{larson2016we}. After examining the fairness metric statistical
parity difference, Angwin et al.~conclude that the algorithm is racially
biased \citep{larson2016we}.

@@

\caption{This plot shows the distribution of two year recidivism outcomes by race. When we divide the data into Black and white defendants, we can see that BLack defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. We can also see that there are more Black defendants in the dataset overall.}\label{fig:recid race plot}
\end{figure}

\begin{longtable}[]{@{}lll@{}}
\caption{This table shows the counts of recidivism by race, illustrating
how a much greater proportion (\textgreater{} 50\%) of Black defendants
recidivated than their white counterparts.
\label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
\midrule
\endhead
African American & 1661 & 1512\tabularnewline
Caucasian & 822 & 1278\tabularnewline
\bottomrule
\end{longtable}

@@

The AIF360 toolkit contains various group and individual fairness
metrics and pre-processing, in-processing, and post-processing
algorithms that we used for our experiments \citep{aif360-oct-2018}. We
researched the definitions and applications of different fairness
metrics \citep{ashokan2021fairness} to determine which metric would be
most appropriate for our project. We chose to look at four group
fairness metrics instead of individual fairness metrics. Group fairness
takes into account the attributes of a whole group as opposed to just
one individual in the group, allowing us to represent more of the
systemic issues happening. In general, group fairness metrics require
that the unprivileged group is treated similarly to the privileged
group, whereas individual fairness metrics require individuals to be
treated consistently \citep{kypraiou_what_2021}. Group and individual
metrics work in opposition of one another, meaning that when group
fairness improves, individual fairness gets worse.

@@

\(P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\)

\textbf{Disparate Impact Ratio} - This metric is the ratio of how often
the favorable outcome occurs in one group versus the other. In the case
of recidivism, this is the ratio of how many white defendants are
predicted to not recidivate compared to how many black defendants are
predicted to not recidivate. A value of 1 means that the ratio is
exactly 1:1. Less than 1 means the privileged group (white defendants)
benefits, while a value greater than 1 means the unprivileged group
(Black defendants) benefits. According to AIF360, a ratio between .8 to
1.25 is considered fair:

\(\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\)

@@

\(\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\)

@@

\(TPR_{D = Unprivileged} - TPR_{D = Privileged}\)

@@

\input{"appendix.tex"}

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{Does the COMPAS Needle Always Point Towards Equity? Finding
Fairness in the COMPAS Risk Assessment Algorithm: A Case Study}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private, for-profit company
Northpointe (now known by its parent company
\href{https://www.equivant.com/faq/}{Equivant}), to predict defendants'
risk of recidivism. It generates a decile score that classifies
defendants' risk of recidivism as either low, medium, or high
\citep{angwin2016machine}. Jurisdictions across the United States use
the COMPAS risk assessment instrument, including but not limited to the
New York \citep{NYDOCCS2019}, Massachusetts, Michigan, California
\citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
Departments of Corrections. Due to the proprietary nature of the COMPAS
algorithm, it is unknown how exactly these recidivism risk scores are
calculated; however, a sample COMPAS Risk Assessment Survey has been
made publicly available, revealing the algorithm's input information.
This survey has been widely critiqued for using proxy variables for race
that do not explicitly factor in a defendant's race but heavily imply
it, allowing Northpointe to claim that their algorithm is free of racial
bias. For example, the COMPAS risk assessment survey asks screeners to
speculate if a defendant might be affiliated with a gang. It also asks
if a defendant has any friends or family members who have been crime
victims \citep{Angwin2016Sample}. Although these questions do not
directly ask about race, they do not take into account the pervasive
nature of systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race. Angwin et
al.~published a 2016 piece in the news publication ProPublica that
analyzes the methods and algorithms used by Northpointe in their COMPAS
risk score assessment algorithm and uncovers the racial biases in
defendants' scores \citep{angwin2016machine}. Their research finds that
``the algorithm {[}is{]} somewhat more accurate than a coin flip,'' a
worrisome level of accuracy given the potential impact its
determinations may have on real people's lives. Angwin et
al.~specifically investigate the distribution of COMPAS scores by decile
among Black and white defendants. They write: ``The analysis also
{[}shows{]} that even when controlling for prior crimes, future
recidivism, age, and gender, Black defendants {[}are{]} 45 percent more
likely to be assigned higher risk scores than white defendants''
\citep{larson2016we}. After examining the fairness metric statistical
parity difference, Angwin et al.~conclude that the algorithm is racially
biased \citep{larson2016we}.

@@

\caption{This plot shows the distribution of two year recidivism outcomes by race. When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. We can also see that there are more Black defendants in the dataset overall.}\label{fig:recid race plot}
\end{figure}

\begin{longtable}[]{@{}lll@{}}
\caption{This table shows the counts of recidivism by race, illustrating
how a much greater proportion (\textgreater{} 50\%) of Black defendants
recidivated than their white counterparts.
\label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived \\
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived \\
\midrule
\endhead
African American & 1661 & 1512 \\
Caucasian & 822 & 1278 \\
\bottomrule
\end{longtable}

@@

The AIF360 toolkit contains various group and individual fairness
metrics as well as pre-processing, in-processing, and post-processing
algorithms that we used for our experiments \citep{aif360-oct-2018}. We
researched the definitions and applications of different fairness
metrics \citep{ashokan2021fairness} to determine which metric would be
most appropriate for our project. We chose to look at four group
fairness metrics instead of individual fairness metrics. Group fairness
takes into account the attributes of a whole group as opposed to just
one individual in the group, allowing us to represent more of the
systemic issues happening. In general, group fairness metrics require
that the unprivileged group is treated similarly to the privileged
group, whereas individual fairness metrics require individuals to be
treated consistently \citep{kypraiou_what_2021}. Group and individual
metrics work in opposition of one another, meaning that when group
fairness improves, individual fairness gets worse.

@@

\[P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\]

\textbf{Disparate Impact Ratio} - This metric is the ratio of how often
the favorable outcome occurs in one group versus the other. In the case
of recidivism, this is the ratio of how many white defendants are
predicted to not recidivate compared to how many Black defendants are
predicted to not recidivate. A value of 1 means that the ratio is
exactly 1:1. Less than 1 means the privileged group (white defendants)
benefits, while a value greater than 1 means the unprivileged group
(Black defendants) benefits. According to AIF360, a ratio between .8 to
1.25 is considered fair:

\[\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\]

@@

\[\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\]

@@

\[TPR_{D = Unprivileged} - TPR_{D = Privileged}\]

@@
